Dynamics of Laterally Propagating Flames in X-ray Bursts. III. Comparing Hotspot Evolution in Two and Three Dimensions

### 2022-12-14 retries/paper.tex

Multiphysics simulation software is hard to implement robustly. One reason is that
physics modules usually have one or more failure modes, and the code may have
additional failure modes related to the coupling between modules. Some examples of
simulation failure modes in our field of astrophysics include:
\begin{itemize}
  \item Numerical integration failures in the system of ordinary differential equations (ODEs) representing nuclear fusion
  \item Numerical integration successes in coupled ODE systems that nevertheless result in nonphysical outcomes, like mass fraction > 1
  \item Convergence failures in iterative linear solvers, such as those used to solve the Poisson equation
  \item Convergence failures in iterative non-linear solvers, such as those used for radiation diffusion
  \item Failures in non-linear root finding, such as in non-monotonic equations of state
  \item Zones whose local velocity exceeds the CFL hydrodynamic stability criterion given the selected timestep
  \item Negative density in a zone after a hydrodynamic update
\end{itemize}
All of these failure modes (and more) are represented in the simulation code Castro \citep{castro_joss}.
In our experience, any useful science simulation will almost inevitably hit one of these failure modes.

@@

A third approach, which we believe is generally more robust and less distortionary, is to retry a timestep
that encounters one of these failure modes. Typically one will use a smaller $\Delta t$ on the retry, although
other options exist, such as modifying the spatial resolution when using an adaptive mesh. One astrophysics
application that successfully does this, which we drew inspiration from for this work, is the stellar evolution
code MESA \citep{MESA}.

>>>>

Multiphysics simulation software is hard to implement robustly. One reason is that
physics modules usually have one or more failure modes, and the code may have
additional failure modes related to the coupling between modules. Some examples of
simulation failure modes in our field of astrophysics include:
\begin{itemize}
  \item Numerical integration failures in the system of ordinary differential equations (ODEs) representing nuclear fusion
  \item Numerical integration successes in coupled ODE systems that nevertheless result in nonphysical outcomes, like mass fraction > 1
  \item Convergence failures in iterative linear solvers, such as those used to solve the Poisson equation
  \item Convergence failures in iterative non-linear solvers, such as those used for radiation diffusion
  \item Failures in non-linear root finding, such as in non-monotonic equations of state
  \item Zones whose local velocity exceeds the CFL hydrodynamic stability criterion given the selected timestep
  \item Negative density in a zone after a hydrodynamic update
\end{itemize}
All of these failure modes (and more) are represented in the simulation code \castro\ \citep{castro_joss}.
In our experience, any useful science simulation will almost inevitably hit one of these failure modes.

@@

A third approach, which we believe is generally more robust and less distortionary, is to retry a timestep
that encounters one of these failure modes. Typically one will use a smaller $\Delta t$ on the retry, although
other options exist, such as modifying the spatial resolution when using an adaptive mesh. One astrophysics
application that successfully does this, which we drew inspiration from for this work, is the stellar evolution
code MESA \citep{MESA}. MESA also has the notion of a ``backup'': a retry can be defined as a second attempt at
the current timestep, while a backup involves returning to an earlier timestep.

\section{Implementation Details}\label{Sec:Implementation}

A functional retry mechanism requires a decision scheme and associated code infrastructure for when to perform
a second attempt at the timestep. A retry scheme effectively turns a timestep into a \texttt{while} loop: attempt
an advance by $\Delta t$, and if it fails then attempt an advance by $\Delta t / 2$, and keep going until you have
reached $t_{i} + \Delta t$ where $t_{i}$ is the starting simulation time. So in both principle and practice,
implementing a retry mechanism is actually quite straightforward, with two complications as noted next.

First, depending on context, this scheme may require saving application state and then restoring that application
state at a later point. For example, \castro\ is a multi-level mesh refinement code where the fine levels are
subcycled with respect to the coarse levels \citep{castro,berger_colella}, and so fine levels typically need to
interpolate coarse data between time levels $t_{i}$ and $t_{i} + \Delta t$, where $\Delta t$ is the coarse grid
timestep. \castro\ does this by maintaining both an ``old'' state and a ``new'' state corresponding to those two
time levels. But subcycling within a given level due to retries will mean that at the end of the coarse step, the
``old'' state data will be at some later time, such as $t_{i} + 3\Delta t / 4$ in the case of two failures. This
needs to be reset back to the initial ``old'' state at $t_{i}$ for valid interpolation onto the fine grid. In \castro\
we handle this by making a backup copy of the data\footnote{For the GPU build of \castro, we save this backup copy
in host pinned memory since we are typically GPU memory constrained} at $t_{i}$ if (and only if) we encounter an
advance failure on the first try.

Second, this scheme may involve refactoring physics modules that perform irrevocable changes to application
state. One way to guarantee that the scheme works is for the sequence of physics modules corresponding to an
advance to be written as an idempotent operator: given an input ``old'' state at $t_{i}$, the advance updates
only the ``new'' state at $t_{i} + \Delta t$, and only in a way that if you re-run the timestep you get the same
result in the ``new'' state. This makes a retry scheme trivial: if you detect a failure, exit from the advance
immediately, and then retry the advance with a smaller timestep.

### 2022-12-14 simplified_sdc_intro/paper.tex

Modeling astrophysical reacting flows can be challenging because of
the disparity between the nuclear and hydrodynamics timescales.
Reaction networks tend to be stiff, requiring implicit integration
techniques to stably integrate the system~\citep{BYRNE19871}.  In contrast, compressible
hydrodynamics flows are limited by the (often much longer)
sound-crossing time over a computational cell and can be solved using
explicit time integration. Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to use different
time-integration methods for the different physics, and to build a
simulation code in a modular way.  However, competition between the
different physical processes can cause the coupling to \replaced{breakdown}{break down, since the reactions do not directly incorporate the effects of the hydrodynamics and vice versa}.
These splitting errors can lead to loss of accuracy and further time
step limitations.

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium. 
These large changes in abundances (and large alternately positive and negative energy generation rates) can be a challenge for
\replaced{a code}{the integration method---it may take more steps than allowed by the ODE integrator, require a timestep below floating point accuracy, or fail to meet the tolerances}.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015,rivas:2022}), but this can make simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and tables for
nuclear statistical equilibrium.

@@

In the Strang splitting~\citep{strang:1968} flavor of operator
splitting, we first integrate the system with reactions terms only (no
advection) over $\Delta t/2$, then integrate the advection terms only
(no reactions) over $\Delta t$, and finally integrate the reaction
terms only over $\Delta t/2$.  \added{The staggering of the reactive update
means the hydro implicitly sees a time-centered reactive source, making
the update second-order accurate in time.}  This flow is illustrated in
Figure~\ref{fig:strang_flowchart}.

@@

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation \replaced{gets}{achieved} second
order convergence and that in particular, integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  
%A variation on Strang splitting called (re-)balanced
%splitting was developed in \citet{speth:2013}, but we do not explore
%that here. \MarginPar{keep this sentence? MK: Don't think it's needed.}

@@

We begin with the general form of a multi-implicit SDC correction equation with two physical processes\deleted{, as originally presented in} \citep{bourlioux2003high}.
Using $(k)$ to denote the iterate, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
In the spirit of \cite{SDC-old}, we treat each advection term as piecewise constant over the time step,
and equal to a time-centered advection term evaluated with the Godunov integrator.
We note that this is a departure from the approach in \cite{bourlioux2003high} that uses a method of lines approach, where 
the advection terms are computed at each node, rather than the midpoint between nodes, using spatial extrapolation only.
Using $\boldsymbol{\mathcal{A}}(\Uc^{n+1/2})$ to denote the advection term, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
Also following \cite{SDC-old}, we differentiate this equation in time
\begin{eqnarray}
\frac{\partial\Uc^{(k)}}{\partial t} = && \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right]\nonumber\\
&+& \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right].
\end{eqnarray}
We combine terms, and arrive at a system of ODEs at each cell,
\begin{equation}
\frac{\partial\Uc^{(k)}}{\partial t} =  \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) \right].\label{eq:sdcupdate}
\end{equation}
We integrate this system from $t^n$ to $t^{n+1}$ using a modified version
of the VODE~\citep{vode} integrator, as described in Section \ref{sec:VODE}.
Thus, we are effectively using two temporal nodes ($t^n$ and $t^{n+1}$) 
and the overall integration is second-order accurate.
In order to properly couple reactions into the computation of the advection terms, 
we use an iteratively lagged reaction source term,
denoted $\Ic_q$ (this is basically an approximation of $\Rb(\qb)$).
The overall flow of the simplified-SDC algorithm is shown in Figure~\ref{fig:sdc_flowchart}.

@@

    \item Predict $\qb$ to the interfaces at $t^{n+1/2}$ using the CTU
      PPM method\added{---this involves taking the cell-centered
        primitive state and tracking all of the information that can
        reach the interface over a timestep, following the procedure
        from \citet{ppmunsplit,millercolella:2002}}.  The source
      terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sc(\qb) = \Gb(\qb) + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      $\Ic_\qb^{n+1/2,(k-1)}$ is a numerical approximation for the effect of
      reactions from the previous iterate, which is computed at the end of each iterate 
      as described below.
      The use of this term contrasts with Strang-splitting, where no
      reactive source terms are included in the hydrodynamics update.
      If $k=1$, we instead use the value from the last iteration of the previous
      time step, $\Ic^{n-1/2,(\kmax)}_{\qb}$.
%%       Any hydrodynamic source terms are time-centered
%%       using the previous iteration:
%% \MarginPar{Formally both $t^n$ or $t^{n+1/2}$ sources give you 2nd-order since they are CTU predictor source terms.  I've never found this type of iteration for this term to amount to any benefit}
%%       \begin{equation}
%%         \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
%%       \end{equation}

      In the unsplit CTU method \citep{ppmunsplit}, the interface
      states used for the final Riemann problem through the zone
      interface consist of a normal predictor and a transverse flux
      correction.  We can add the source terms either to the normal
      predictor (for example, doing characteristic tracing as
      described in \citealt{ppm}) or after all of the transverse flux
      corrections are made.  Both are second-order accurate.  For the
      gravitational sources, we do those in the normal predictor\added{, consistent
      with the formulation in \citet{ppm}.}
      However, for the reactive sources, we found that it is most
      reliable to add them at the end of the interface state
      construction, after the transverse flux corrections.  This is
      because we want to ensure that sum over species of $\Ic_\qb$ is
      zero, and characteristic tracing or the various flux corrections
      in general do not preserve this.  We enforce that the species interface
      states remain in $[0, 1]$ after adding the reactive source.

@@

    Our integrator also needs the Jacobian of the system, in terms of
    $\Uc^\prime$.  This is different than the form of the Jacobian
    usually used in reaction networks \added{(we depend on $e$ instead of
    $T$)}.  We describe the form of the Jacobian in Appendix
    \ref{sec:app:jacobian}.

@@

We use a modified version of the VODE~\citep{vode} integrator,
designed for stiff ODEs.  Our version has been ported to \replaced{C++, as a
header-only library enabling it to run on GPUs}{modern C++ and designed to run on GPUs}.  We have also added
checks to the timestep rejection logic that help prevent VODE from
wandering too far off the solution\footnote{This modified version of
VODE is available in our Microphysics repo:
\url{https://github.com/AMReX-Astro/Microphysics}.}.  We have found empirically that these checks improve the performance of VODE.

@@

Tables \ref{table:react_converge_strang} and
\ref{table:react_converge_sdc} show the Strang splitting and
simplified-SDC convergence respectively.  \added{We ran each integration method on a domain with $64^2$, $128^2$, $256^2$, $512^2$ zones.  An error is defined between successive resolutions by coarsening the higher resolution run and taking the L1 norm of the zone-by-zone difference of the two simulations---for example, $\epsilon_{64\rightarrow 128}$ is the error between the $64^2$ and $128^2$ simulations.}  \deleted{The convergence rate is
computed by taking log-base-2 of the ratio of adjacent errors.}  \added{The convergence
rate from $\epsilon_{64\rightarrow 128}$ to $\epsilon_{128\rightarrow 256}$ is computed as $\log_2 \epsilon_{64\rightarrow 128}/\epsilon_{128\rightarrow 256}$ and is given in the column between those errors, and likewise for the higher-resolution case.}  A rate
of 2.0 indicates second order convergence.  We see that both methods
\replaced{converge nearly second order, and agree well with each
  other}{exhibit convergence at roughly second-order and agree well.}.
This demonstrates that both methods are working as designed.

@@

Figure \ref{fig:subch_strang_cfl02} shows the temperature, mean
molecular weight, and nuclear energy generation rate for the Strang splitting
CFL = 0.2 simulation ({\tt strang\_subch2\_cfl0.2}).  At this point
in the evolution, the helium detonation has wrapped more than halfway
across the surface and the ingoing carbon detonation is approaching the
center of the star.  We see that the
mean molecular weight appears mottled behind the inward propagating
detonation in the white dwarf.  Comparing to the energy generation
rate, we see that region is characterized by large energy releases of
alternating signs checkerboarding thoughout the region.  This is
characteristic of the nucleosynthesis struggling to come into
equilibrium at high temperatures when the reverse rates can be
important.  \added{To highlight this, Figure~\ref{fig:strang_steps} shows the nuclear
energy generation rate in a highly-zoomed region around the detonation over four
consecutive coarse timesteps.  We see that on the detonation front itself, the
sign of the energy generation rate changes from one step to the next in many zones.}

>>>>

Modeling astrophysical reacting flows can be challenging because of
the disparity between the nuclear and hydrodynamics timescales.
Reaction networks tend to be stiff, requiring implicit integration
techniques to stably integrate the system~\citep{BYRNE19871}.  In contrast, compressible
hydrodynamics flows are limited by the (often much longer)
sound-crossing time over a computational cell and can be solved using
explicit time integration. Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to use different
time-integration methods for the different physics, and to build a
simulation code in a modular way.  However, competition between the
different physical processes can cause the coupling to \edit1{break down, since the reactions do not directly incorporate the effects of the hydrodynamics and vice versa}.
These splitting errors can lead to loss of accuracy and further time
step limitations.

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium. 
These large changes in abundances (and large alternately positive and negative energy generation rates) can be a challenge for
\edit1{the integration method---it may take more steps than allowed by the ODE integrator, require a timestep below floating point accuracy, or fail to meet the tolerances}.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015,rivas:2022}), but this can make simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and tables for
nuclear statistical equilibrium.

@@

In the Strang splitting~\citep{strang:1968} flavor of operator
splitting, we first integrate the system with reactions terms only (no
advection) over $\Delta t/2$, then integrate the advection terms only
(no reactions) over $\Delta t$, and finally integrate the reaction
terms only over $\Delta t/2$.  \edit1{The staggering of the reactive update
means the hydro implicitly sees a time-centered reactive source, making
the update second-order accurate in time.}  This flow is illustrated in
Figure~\ref{fig:strang_flowchart}.

@@

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation \edit1{achieved} second
order convergence and that in particular, integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  
%A variation on Strang splitting called (re-)balanced
%splitting was developed in \citet{speth:2013}, but we do not explore
%that here. \MarginPar{keep this sentence? MK: Don't think it's needed.}

@@

We begin with the general form of a multi-implicit SDC correction equation with two physical processes.
Using $(k)$ to denote the iterate, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
In the spirit of \cite{SDC-old}, we treat each advection term as piecewise constant over the time step,
and equal to a time-centered advection term evaluated with the Godunov integrator.
We note that this is a departure from the approach in \cite{bourlioux2003high} that uses a method of lines approach, where 
the advection terms are computed at each node, rather than the midpoint between nodes, using spatial extrapolation only.
Using $\boldsymbol{\mathcal{A}}(\Uc^{n+1/2})$ to denote the advection term, we have
\begin{eqnarray}
\Uc^{(k)}(t) = \Uc^n &+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right ] dt \nonumber\\
&+& \int_{t^n}^t \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right ] dt.
\end{eqnarray}
Also following \cite{SDC-old}, we differentiate this equation in time
\begin{eqnarray}
\frac{\partial\Uc^{(k)}}{\partial t} = && \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) - \Advs{\Uc^{n+1/2,(k-1)}} - \Rb\left(\Uc^{(k-1)}\right) \right]\nonumber\\
&+& \left [ \Advs{\Uc^{n+1/2,(k-1)}} + \Rb\left(\Uc^{(k-1)}\right) \right].
\end{eqnarray}
We combine terms, and arrive at a system of ODEs at each cell,
\begin{equation}
\frac{\partial\Uc^{(k)}}{\partial t} =  \left [ \Advs{\Uc^{n+1/2,(k)}} + \Rb\left(\Uc^{(k)}\right) \right].\label{eq:sdcupdate}
\end{equation}
We integrate this system from $t^n$ to $t^{n+1}$ using a modified version
of the VODE~\citep{vode} integrator, as described in Section \ref{sec:VODE}.
Thus, we are effectively using two temporal nodes ($t^n$ and $t^{n+1}$) 
and the overall integration is second-order accurate.
In order to properly couple reactions into the computation of the advection terms, 
we use an iteratively lagged reaction source term,
denoted $\Ic_q$ (this is basically an approximation of $\Rb(\qb)$).
The overall flow of the simplified-SDC algorithm is shown in Figure~\ref{fig:sdc_flowchart}.

@@

    \item Predict $\qb$ to the interfaces at $t^{n+1/2}$ using the CTU
      PPM method\edit1{---this involves taking the cell-centered
        primitive state and tracking all of the information that can
        reach the interface over a timestep, following the procedure
        from \citet{ppmunsplit,millercolella:2002}}.  The source
      terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sc(\qb) = \Gb(\qb) + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      $\Ic_\qb^{n+1/2,(k-1)}$ is a numerical approximation for the effect of
      reactions from the previous iterate, which is computed at the end of each iterate 
      as described below.
      The use of this term contrasts with Strang-splitting, where no
      reactive source terms are included in the hydrodynamics update.
      If $k=1$, we instead use the value from the last iteration of the previous
      time step, $\Ic^{n-1/2,(\kmax)}_{\qb}$.
%%       Any hydrodynamic source terms are time-centered
%%       using the previous iteration:
%% \MarginPar{Formally both $t^n$ or $t^{n+1/2}$ sources give you 2nd-order since they are CTU predictor source terms.  I've never found this type of iteration for this term to amount to any benefit}
%%       \begin{equation}
%%         \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
%%       \end{equation}

      In the unsplit CTU method \citep{ppmunsplit}, the interface
      states used for the final Riemann problem through the zone
      interface consist of a normal predictor and a transverse flux
      correction.  We can add the source terms either to the normal
      predictor (for example, doing characteristic tracing as
      described in \citealt{ppm}) or after all of the transverse flux
      corrections are made.  Both are second-order accurate.  For the
      gravitational sources, we do those in the normal predictor\edit1{, consistent
      with the formulation in \citet{ppm}.}
      However, for the reactive sources, we found that it is most
      reliable to add them at the end of the interface state
      construction, after the transverse flux corrections.  This is
      because we want to ensure that sum over species of $\Ic_\qb$ is
      zero, and characteristic tracing or the various flux corrections
      in general do not preserve this.  We enforce that the species interface
      states remain in $[0, 1]$ after adding the reactive source.

@@

    Our integrator also needs the Jacobian of the system, in terms of
    $\Uc^\prime$.  This is different than the form of the Jacobian
    usually used in reaction networks \edit1{(we depend on $e$ instead of
    $T$)}.  We describe the form of the Jacobian in Appendix
    \ref{sec:app:jacobian}.

@@

We use a modified version of the VODE~\citep{vode} integrator,
designed for stiff ODEs.  Our version has been ported to \edit1{modern C++ and designed to run on GPUs}.  We have also added
checks to the timestep rejection logic that help prevent VODE from
wandering too far off the solution\footnote{This modified version of
VODE is available in our Microphysics repo:
\url{https://github.com/AMReX-Astro/Microphysics}.}.  We have found empirically that these checks improve the performance of VODE.

@@

Tables \ref{table:react_converge_strang} and
\ref{table:react_converge_sdc} show the Strang splitting and
simplified-SDC convergence respectively.  \edit1{We ran each integration method on a domain with $64^2$, $128^2$, $256^2$, $512^2$ zones.  An error is defined between successive resolutions by coarsening the higher resolution run and taking the L1 norm of the zone-by-zone difference of the two simulations---for example, $\epsilon_{64\rightarrow 128}$ is the error between the $64^2$ and $128^2$ simulations.}  \edit1{The convergence
rate from $\epsilon_{64\rightarrow 128}$ to $\epsilon_{128\rightarrow 256}$ is computed as $\log_2 \epsilon_{64\rightarrow 128}/\epsilon_{128\rightarrow 256}$ and is given in the column between those errors, and likewise for the higher-resolution case.}  A rate
of 2.0 indicates second order convergence.  We see that both methods
\edit1{exhibit convergence at roughly second-order and agree well.}.
This demonstrates that both methods are working as designed.

@@

Figure \ref{fig:subch_strang_cfl02} shows the temperature, mean
molecular weight, and nuclear energy generation rate for the Strang splitting
CFL = 0.2 simulation ({\tt strang\_subch2\_cfl0.2}).  At this point
in the evolution, the helium detonation has wrapped more than halfway
across the surface and the ingoing carbon detonation is approaching the
center of the star.  We see that the
mean molecular weight appears mottled behind the inward propagating
detonation in the white dwarf.  Comparing to the energy generation
rate, we see that region is characterized by large energy releases of
alternating signs checkerboarding thoughout the region.  This is
characteristic of the nucleosynthesis struggling to come into
equilibrium at high temperatures when the reverse rates can be
important.  \edit1{To highlight this, Figure~\ref{fig:strang_steps} shows the nuclear
energy generation rate in a highly-zoomed region around the detonation over four
consecutive coarse timesteps.  We see that on the detonation front itself, the
sign of the energy generation rate changes from one step to the next in many zones.}

### 2022-12-14 xrb3/paper.tex

@@

%% \author[0000-0001-6191-4285]{Kiran Eiden}
%% \affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
%%              Stony Brook, NY 11794-3800}

%% \author[0000-0001-8401-030X]{Michael Zingale}
%% \affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
%%              Stony Brook, NY 11794-3800}

@@

%% \author{Yuri Cavecchi}
%% \affiliation{Mathematical Sciences and STAG Research Centre, University of Southampton, SO17 1BJ}

%% \author[0000-0003-0439-4556]{Max P.\ Katz}
%% \affiliation{NVIDIA Corp}

%% \author[0000-0001-8092-1974]{Weiqun Zhang}
%% \affiliation{Lawrence Berkeley National Laboratory, Berkeley, CA}

%% \correspondingauthor{Alice Harpole}
%% \email{alice.harpolee@stonybrook.edu}

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We continue to investigate laterally propagating flames in XRBs,
comparing the early evolution and structure of a burning hotspot
between two- and three-dimensions.
\end{abstract}

@@

X-ray bursts result from thermonuclear burning of an accreted H/He or
He layer on a neutron star \citep{galloway:2017}.  Observations of
brightness oscillations in the rise of the lightcurve suggest that the
burning begins localized and then spreads across the neutron star.
One-dimensional simulations of X-ray bursts reproduce the lightcurve
and recurrence time \citep{woosley-xrb} and can use large reaction
networks to explore rp-process nucleosynthesis.

@@

We have been performing simulations of flame spreading on a neutron
star resolving the flame structure using realistic reaction networks.  Our two-dimensional studies \citep{eiden:2020,harpole:2021} demonstrated our open-source framework for modeling X-ray bursts using

We use the \castro\ \citep{castro,castro_joss} simulation code, following
the same numerical approach described in \citet{eiden:2020,harpole:2021}.

>>>>

\author[0000-0001-8401-030X]{Michael Zingale}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}

@@

\author[0000-0001-6191-4285]{Kiran Eiden}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}

@@

\author[0000-0003-0439-4556]{Max P.\ Katz}
\affiliation{NVIDIA Corp}

\correspondingauthor{Michael Zingale}
\email{michael.zingale@stonybrook.edu}

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We explore the early evolution of a hot spot and a flame ignition and
spreading on the surface of a neutron star in three-dimensions.  We
focus on the difference in the structure of the flame compared to
previous two-dimensional simulations.\end{abstract}

@@

X-ray bursts (XRBs) result from thermonuclear burning of an accreted
H/He or He layer on a neutron star \citep{galloway:2017}.
Observations of brightness oscillations in the rise of the lightcurve
suggest that the burning begins localized and then spreads across the
neutron star.  This spreading is inherently a multi-dimensional
phenomena, and hydrodynamic simulations that resolve the reactive zone
and capture the scales over which rotation is important are needed to
understand the flame propagation.  We show a first attempt at modeling
the full three-dimensional ignition and spreading of a hot spot in an
XRB.  This builds on our earlier two dimensional work
\citep{eiden:2020,harpole:2021} that developed our simulation
framework and explored how the acceleration of the burning front
depends on the initial model structure.  As with those simulations, we
use the freely-available \castro\ simulation code
\citep{castro,castro_joss} simulation code.

@@

The goal of this paper is to understand the challenges of a fully-resolved
three-dimensional simulation and explore what is possible.  This will inform
our follow-on simulations.
